---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

# XGeoRTR
## Frédéric Bertrand

`XGeoRTR` is explainable geometry backend infrastructure for R, positioned for
SIGGRAPH 2026 workflows that need reusable state rather than package-specific
presentation code. It owns geometry-aware state, embeddings, diagnostics,
multiscale summaries, selection state, selected backend tables, and JSON state
exchange.

Downstream packages such as `shapViz3D`, `rTDA3D`, and renderer frontends can
consume `xgeo_state` or exported backend tables without XGeoRTR taking ownership
of their display, interaction, or use-case presentation layers.

## Role in GeoXGL

`XGeoRTR` is the geometry/state backend of the [`GeoXGL`](https://fbertran.github.io/GeoXGL/)
ecosystem, a modular R/WebGL stack for scalable statistical graphics,
explanation geometry, and scientific visualization.

Within GeoXGL:

- `XGeoRTR` owns backend-neutral explanation geometry, `xgeo_state`, embeddings,
  diagnostics, level-of-detail summaries, selection state, public backend tables,
  and JSON state exchange.
- `ggWebGL` owns browser-native WebGL rendering, widgets, viewport interaction,
  shader execution, and static export surfaces.
- `shapViz3D` and other use-case packages own domain semantics and downstream
  layout mappings.

This package does not own rendering, viewport orchestration, camera control,
canvas management, shader execution, widget construction, visual themes, or
export surfaces. Those responsibilities remain in renderer frontends such as
`ggWebGL`.

## Current scope and future work

**Current scope:** backend state, coercion, embeddings, diagnostics,
level-of-detail summaries, selections, explanation tables, point-value tables,
regular-grid tables, and JSON serialization.

**Future/prototype scope:** topology-specific reasoning, prediction-regime path
semantics, and large-scale performance claims should be presented only when
supported by concrete downstream examples or benchmark artifacts. Causal
explanation is outside the current package scope.

## Main features

- Canonical backend object: `xgeo_state()`
- Coercion: `as_xgeo_state()`
- State operators:
  - `compute_xgeo_embedding()`
  - `compute_xgeo_diagnostics()`
  - `build_xgeo_lod()`
  - `set_active_embedding()`
  - `set_xgeo_selection()`
  - `set_xgeo_lod()`
- Backend-neutral accessors:
  - `xgeo_geometry()`
  - `xgeo_attributes()`
  - `xgeo_indices()`
  - `xgeo_selection()`
  - `xgeo_metadata()`
- Backend-neutral tables:
  - `xgeo_explanation_table()`
  - `xgeo_point_values()`
  - `xgeo_regular_grid()`
- Serialization:
  - `write_xgeo_state()`
  - `read_xgeo_state()`

## Installation

For local development from checked-out repositories:

```r
install.packages(c("cli", "jsonlite"))
devtools::load_all(".")
```

Optional packages:

- `uwot` for optional UMAP embeddings
- `knitr`, `rmarkdown`, and `pkgdown` for documentation work

## Build backend state

```{r state-workflow, eval = FALSE}
library(XGeoRTR)

demo_path <- system.file("extdata", "spatial_demo.csv", package = "XGeoRTR")
demo_tbl <- utils::read.csv(demo_path, stringsAsFactors = FALSE)

state <- as_xgeo_state(
  demo_tbl,
  x_col = "x",
  y_col = "y",
  z_col = "z",
  value_col = "value",
  feature_col = "feature",
  method = "spatial-field-demo",
  meta = list(source = "synthetic-demo", sample_id = "grid-01")
)

state <- compute_xgeo_embedding(state, method = "pca", source = "points", dims = 2)
state <- set_active_embedding(state, "pca_points")
state <- compute_xgeo_diagnostics(state, embedding = "pca_points", source = "points", k = 3)
state <- build_xgeo_lod(state, embedding = "pca_points", levels = c(8L, 16L), auto_threshold = 10L)
state <- set_xgeo_selection(state, point_ids = state$indices$point_ids[[1]])

summary(state)
```

## Write and reload state

```{r state-io, eval = FALSE}
out_json <- tempfile(fileext = ".json")
write_xgeo_state(state, out_json)
restored_state <- read_xgeo_state(out_json)
```

## Build downstream tables

```{r state-tables, eval = FALSE}
long_tbl <- xgeo_explanation_table(state)
point_tbl <- xgeo_point_values(state)
grid <- xgeo_regular_grid(point_tbl)
```

Use-case packages should consume these public tables instead of inspecting
internal ingestion objects. For example, a Shapley-oriented package can consume
selected explanation tables, while a topology-oriented package can consume point
and regular-grid summaries as backend inputs.

## Supported poster claims

The package directly supports these backend claims:

- canonical backend state: `xgeo_state`
- tabular-to-state ingestion: `as_xgeo_state`
- embedding and diagnostic operators for geometry-aware workflows
- multiscale LOD summaries with `build_xgeo_lod`
- selected explanation, point-value, and regular-grid tables
- backend JSON exchange with `write_xgeo_state` and `read_xgeo_state`
- downstream-consumer readiness for packages such as `shapViz3D`, `rTDA3D`,
  and renderer frontends

## Downstream figure consumers

`XGeoRTR` does not ship the canonical selected poster figures for SHAP
workflows. Those assets and their SHAP semantics live in the `shapViz3D`
repository under `siggraph_figures/final_selection/`.

What `XGeoRTR` provides is the backend path that those downstream figures
consume:

- `as_xgeo_state()` standardizes the long tables
- `set_xgeo_selection()` controls subset state upstream
- `xgeo_explanation_table()`, `xgeo_point_values()`, and `xgeo_regular_grid()`
  expose renderer-neutral tables downstream
- `compute_xgeo_embedding()`, `compute_xgeo_diagnostics()`, and
  `build_xgeo_lod()` provide optional backend context without taking over
  presentation

The backend-only example below shows how the three `shapViz3D` evidence tables
enter the public `xgeo_state` pipeline when the sibling `shapViz3D` repo is
available. If that repo is unavailable, the same example falls back to the
bundled `spatial_demo.csv` so the backend workflow still runs without a SHAP or
renderer dependency.

```{r eval = FALSE}
source("inst/examples/downstream_shapviz3d_state_tables.R")
```

## Scope

`XGeoRTR` produces backend state and computation products; downstream packages
decide how to present or interact with those products.

- XGeoRTR: backend geometry/state semantics, operators, selected tables, and
  serialization
- Downstream consumers: Shapley presentation, TDA workflows, renderer adapters,
  and front-end interaction

The frozen package boundary is documented in `INTERFACE_FREEZE.md`. Public
`XGeoRTR` objects and serialized payloads must remain free of display-system
fields such as scene, layer, viewport, camera, canvas, theme, shader, widget,
and export surface metadata.
