---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# XGeoRTR
## Frédéric Bertrand

`XGeoRTR` is the general platform package for spatial explanation geometry in R.
It owns the neutral data model, scene abstraction, regular-grid surface view,
and `rgl` / WebGL export path that can support SHAP and non-SHAP workflows.

## Main features

- `xgeo_data` S3 objects for generic spatial explanation tables.
- `as_xgeo_data()` helpers that standardize matrices and long tables into a
  common scene-ready structure.
- `xgeo_scene()` plus `geom_xgeo_surface()` for a generic spatial surface view.
- `render_webgl()` for rendering to `rgl` and exporting HTML widgets when
  `htmlwidgets` is available.
- A packaged spatial demo dataset under `inst/extdata/`.

## Installation

For local development from checked-out repositories:

```r
install.packages(c("cli", "rgl"))
devtools::load_all(".")
```

Optional packages:

- `htmlwidgets` for `render_webgl(file = ...)`
- `knitr`, `rmarkdown`, and `pkgdown` for documentation work

## A first generic workflow

```{r first-workflow, eval = FALSE}
library(XGeoRTR)

demo_path <- system.file("extdata", "spatial_demo.csv", package = "XGeoRTR")
demo_tbl <- utils::read.csv(demo_path, stringsAsFactors = FALSE)

xd <- as_xgeo_data(
  demo_tbl,
  x_col = "x",
  y_col = "y",
  z_col = "z",
  value_col = "value",
  feature_col = "feature",
  method = "surface-demo",
  meta = list(source = "synthetic-demo", sample_id = "grid-01")
)

summary(xd)
```

## Render a generic surface

```{r render-surface, eval = FALSE}
scene <- xgeo_scene(xd, camera = list(preset = "top")) +
  geom_xgeo_surface(alpha = 0.7, smooth = TRUE)

render_webgl(scene)
```

## Status

`XGeoRTR` is the general platform layer for the current spatial-only MVP. In
this first wave it intentionally covers only:

- generic precomputed explanation tables
- one spatial structure
- one working `rgl` renderer
- one generic surface layer
- HTML export for package and poster workflows
