---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# XGeoRTR
## FrĂ©dĂ©ric Bertrand

`XGeoRTR` is the general platform package for explanation geometry in R. It
owns the neutral data model, scene abstraction, embeddings, diagnostics,
multiscale summaries, and `rgl` / WebGL export path.

## Main features

- Normalized `xgeo_data` objects with explicit `points`, `explanations`,
  `point_meta`, `feature_meta`, `predictions`, and `uncertainty`.
- `as_xgeo_data()` preserves unmapped point-level metadata during ingestion.
- `xgeo_scene()` owns embeddings, diagnostics, selection state, LOD state, and
  scene metadata.
- `compute_xgeo_embedding()`, `compute_xgeo_diagnostics()`, and
  `build_xgeo_lod()` provide a first platform path for embeddings,
  trustworthiness, local agreement, and density-grid multiscale summaries.
- `geom_xgeo_surface()`, `geom_xgeo_points()`, and `geom_xgeo_density()` cover
  regular-grid, point-cloud, and density views.
- `render_webgl()` renders scenes to `rgl` and exports HTML widgets when
  `htmlwidgets` is available.
- `write_xgeo_scene()` / `read_xgeo_scene()` provide JSON scene interchange.

## Installation

For local development from checked-out repositories:

```r
install.packages(c("cli", "jsonlite", "rgl"))
devtools::load_all(".")
```

Optional packages:

- `htmlwidgets` for `render_webgl(file = ...)`
- `uwot` for optional UMAP embeddings
- `knitr`, `rmarkdown`, and `pkgdown` for documentation work

## A first generic workflow

```{r first-workflow, eval = FALSE}
library(XGeoRTR)

demo_path <- system.file("extdata", "spatial_demo.csv", package = "XGeoRTR")
demo_tbl <- utils::read.csv(demo_path, stringsAsFactors = FALSE)

xd <- as_xgeo_data(
  demo_tbl,
  x_col = "x",
  y_col = "y",
  z_col = "z",
  value_col = "value",
  feature_col = "feature",
  method = "surface-demo",
  meta = list(source = "synthetic-demo", sample_id = "grid-01")
)

summary(xd)
```

## Build a platform scene

```{r platform-scene, eval = FALSE}
scene <- xgeo_scene(xd, camera = list(preset = "top"))
scene <- compute_xgeo_embedding(scene, method = "pca", source = "points", dims = 2)
scene <- set_active_embedding(scene, "pca_points")
scene <- compute_xgeo_diagnostics(scene, embedding = "pca_points", source = "points", k = 3)
scene <- build_xgeo_lod(scene, embedding = "pca_points", levels = c(8L, 16L), auto_threshold = 10L)
```

## Render point and density views

```{r render-views, eval = FALSE}
point_scene <- scene + geom_xgeo_points(color_by = "local_agreement", size = 8)
density_scene <- scene + geom_xgeo_density(lod_name = "density_grid_pca_points", level = "16")
surface_scene <- xgeo_scene(xd, camera = list(preset = "top")) +
  geom_xgeo_surface(alpha = 0.7, smooth = TRUE)

render_webgl(point_scene, lod_level = "auto")
render_webgl(density_scene)
render_webgl(surface_scene)
```

## Write and reload a scene

```{r scene-io, eval = FALSE}
out_json <- tempfile(fileext = ".json")
write_xgeo_scene(point_scene, out_json)
restored_scene <- read_xgeo_scene(out_json)
```

## Status

`XGeoRTR` is the general platform layer for the current spatial-only release.
It already owns:

- scene model
- embeddings
- diagnostics
- density-grid LOD summaries
- interaction state through explicit selection and active-view state
- JSON export/import
- point, density, and surface rendering

The current boundary is still deliberate:

- one structure type: `spatial`
- one rendering backend: `rgl`
- no browser-side zoom callbacks yet
- no mesh or event-time renderers yet
