---
title: "Getting Started with XGeoRTR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with XGeoRTR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rgl.useNULL = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`XGeoRTR` is the general platform package for the current spatial-only MVP. It
owns the neutral explanation object, scene abstraction, generic surface layer,
and `rgl` / WebGL export path.

The first release is intentionally narrow:

- one neutral data object: `xgeo_data`
- one composable scene container: `xgeo_scene()`
- one generic regular-grid view: `geom_xgeo_surface()`
- one export path: `render_webgl()`

## Create an `xgeo_data` object

```{r}
library(XGeoRTR)

demo_path <- system.file("extdata", "spatial_demo.csv", package = "XGeoRTR")
demo_tbl <- utils::read.csv(demo_path, stringsAsFactors = FALSE)

xd <- as_xgeo_data(
  demo_tbl,
  x_col = "x",
  y_col = "y",
  z_col = "z",
  value_col = "value",
  feature_col = "feature",
  method = "surface-demo",
  meta = list(source = "synthetic-demo", sample_id = "grid-01")
)

xd
summary(xd)
```

## Build a scene

```{r}
scene <- xgeo_scene(xd, camera = list(preset = "top")) +
  geom_xgeo_surface(alpha = 0.7, smooth = TRUE)

scene
```

## Render to WebGL

Rendering requires `rgl`. Saving to an HTML file also requires `htmlwidgets`.

```{r, eval = FALSE}
render_webgl(scene)

outfile <- tempfile(fileext = ".html")
render_webgl(scene, file = outfile, selfcontained = FALSE)
```

## Platform position

`XGeoRTR` is the substrate for explanation packages rather than a SHAP package
itself. SHAP-specific semantics and layouts now belong in `shapViz3D`, while
`XGeoRTR` owns the parts that still make sense for non-SHAP explanation tables:

- generic ingestion
- scene composition
- regular-grid rendering
- camera/orbit metadata
- HTML export
