---
title: "Getting Started with XGeoRTR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with XGeoRTR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
options(rgl.useNULL = TRUE)

knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

`XGeoRTR` is the general platform package for the current spatial-only release.
It owns the neutral explanation object, scene abstraction, embeddings,
diagnostics, multiscale summaries, selection state, and `rgl` / WebGL export
path.

The current release includes:

- one neutral data object: `xgeo_data`
- one composable scene container: `xgeo_scene()`
- one embedding path: `compute_xgeo_embedding()`
- one diagnostic path: `compute_xgeo_diagnostics()`
- one density-grid LOD path: `build_xgeo_lod()`
- three generic views: `geom_xgeo_surface()`, `geom_xgeo_points()`, and
  `geom_xgeo_density()`
- one export path: `render_webgl()` plus JSON scene IO

## Create an `xgeo_data` object

```{r}
library(XGeoRTR)

demo_path <- system.file("extdata", "spatial_demo.csv", package = "XGeoRTR")
demo_tbl <- utils::read.csv(demo_path, stringsAsFactors = FALSE)

xd <- as_xgeo_data(
  demo_tbl,
  x_col = "x",
  y_col = "y",
  z_col = "z",
  value_col = "value",
  feature_col = "feature",
  method = "surface-demo",
  meta = list(source = "synthetic-demo", sample_id = "grid-01")
)

xd
summary(xd)
```

## Build a scene and attach platform state

```{r}
scene <- xgeo_scene(xd, camera = list(preset = "top"))
scene <- compute_xgeo_embedding(scene, method = "pca", source = "points", dims = 2)
scene <- set_active_embedding(scene, "pca_points")
scene <- compute_xgeo_diagnostics(scene, embedding = "pca_points", source = "points", k = 3)
scene <- build_xgeo_lod(scene, embedding = "pca_points", levels = c(8L, 16L), auto_threshold = 10L)

scene
```

## Render point, density, and surface views

Rendering requires `rgl`. Saving to an HTML file also requires `htmlwidgets`.

```{r, eval = FALSE}
point_scene <- scene + geom_xgeo_points(color_by = "local_agreement", size = 8)
density_scene <- scene + geom_xgeo_density(lod_name = "density_grid_pca_points", level = "16")
surface_scene <- xgeo_scene(xd, camera = list(preset = "top")) +
  geom_xgeo_surface(alpha = 0.7, smooth = TRUE)

render_webgl(point_scene, lod_level = "auto")
render_webgl(density_scene)
render_webgl(surface_scene)

outfile <- tempfile(fileext = ".html")
render_webgl(point_scene, file = outfile, selfcontained = FALSE)
```

## Write and reload scene state

```{r, eval = FALSE}
json_file <- tempfile(fileext = ".json")
write_xgeo_scene(point_scene, json_file)
restored <- read_xgeo_scene(json_file)
```

## Platform position

`XGeoRTR` is the substrate for explanation packages: it owns the parts that
still make sense outside SHAP semantics:

- generic ingestion with metadata preservation
- scene composition and explicit selection state
- embeddings and diagnostics
- multiscale density summaries
- generic point, density, and surface rendering
- HTML and JSON export/import

SHAP-specific semantics and layouts continue to belong in `shapViz3D`.
